Technique for \earlywarning\ detection of compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-14 introduction.tex
removed algorithmic latency
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1018 ccdd05d5-56c1-461a-bf98-85f6876523b4
minor edits
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1073 ccdd05d5-56c1-461a-bf98-85f6876523b4

In the era of first-generation detectors, the \GW{} community initiated a
project to send alerts when potential \GW{} transients were observed in
order to trigger followup observations by \EM{} telescopes.  The best
demonstrated latencies were on the order of 1 hour%
~\citep{HugheyGWPAW2011}, which was an important
achievement, but far too late to catch prompt \EM{}
emission. In order to extract the maximum science from Advanced \LIGO{}
\CBC{} events, we have the ambition of reporting \GW{} candidates not minutes
\emph{after} the merger, as has already been accomplished, but seconds
\emph{before}.

@@

We show that by examining the signal to noise ratio (\SNR{}) stream for
threshold crossings before the \GW{} signal leaves the detection band, it is
possible to trade some \SNR{} and sky localization
accuracy for negative latency.  Figure~\ref{fig:earlywarning} shows
projected early detection rates for \textsc{ns}--\textsc{ns} binaries in Advanced
\LIGO\ assuming event rates predicted in \citet{Abadie:2010p10836} and anticipated
detector sensitivity for the `zero detuning, high power' configuration described in 
\citet{ALIGONoise}.  The most realistic estimates indicate that at an \SNR\ threshold
of 8, we will observe a total of 40~events~yr$^{-1}$, but $\sim$10~yr$^{-1}$ will be
detectable within 10~s of merger and $\sim$1~yr$^{-1}$ will be detectable within
100~s of merger.  These rates are uncertain by an order of magnitude.
%
\begin{figure}[h]
\plotone{figures/snr_in_time.pdf}
\caption{\label{fig:earlywarning}Expected number of \textsc{ns}--\textsc{ns}
sources that will be detectable by Advanced \LIGO\ $t$ seconds before coalescence.
The heavy solid line is the most realistic yearly rate estimate.  The shaded
region represents the 5 to 95\% confidence interval arising from uncertainty
in predicted event rates \citep{Abadie:2010p10836}.}
\end{figure}

@@

Realizing advance detection of compact binary coalescences (\CBC{}s) will
require striking a balance between latency and throughput. \CBC{} searches
consist of banks of matched filters, or
cross-correlations between the data stream and a bank of nominal ``template''
signals.  There are many different implementations of matched filters, but most
have high throughput at the cost of high latency, or low latency at the cost of
low throughput.  The former are epitomized by the overlap-save algorithm~%
\citep{numerical-recipes-chapter-13} for frequency domain (\FD) convolution,
currently the preferred method in \GW{}
searches.  The most obvious example of the latter is the time domain
(\TD) convolution, which has no algorithmic latency.  However, its
computational complexity is quadratic in the length of the templates, so it is
prohibitively expensive for long templates.

>>>>

In the era of first-generation detectors, the \GW{} community initiated a
project to send alerts when potential \GW{} transients were observed in order
to trigger followup observations by \EM{} telescopes.  The typical latencies
were 30--60 minutes ~\citep{HugheyGWPAW2011}, which was an important
achievement, but far too late to catch prompt \EM{} emission. We have the
ambition of reporting \GW{} candidates not minutes \emph{after} the merger, as
has already been accomplished, but seconds \emph{before}.

@@

We show that by examining the signal to noise ratio (\SNR{}) stream for
threshold crossings before the \GW{} signal leaves the detection band, it is
possible to trade some \SNR{} and sky localization accuracy for pre-merger
detection.  Figure~\ref{fig:earlywarning} shows projected early detection rates
for \textsc{ns}--\textsc{ns} binaries in Advanced \LIGO\ assuming event rates
predicted in \citet{Abadie:2010p10836} and anticipated detector sensitivity for
the `zero detuning, high power' configuration described in \citet{ALIGONoise}.
The most realistic estimates indicate that at an \SNR\ threshold of 8, we will
observe a total of 40~events~yr$^{-1}$, but $\sim$10~yr$^{-1}$ will be
detectable within 10~s of merger and $\sim$1~yr$^{-1}$ will be detectable
within 100~s of merger.  These rates are uncertain and the 5 to 95\% confidence
interval is shown in figure \ref{fig:earlywarning}.
%
\begin{figure}[h]
\plotone{figures/snr_in_time.pdf}
\caption{\label{fig:earlywarning}Expected number of \textsc{ns}--\textsc{ns}
sources that will be detectable by Advanced \LIGO\ $t$ seconds before coalescence.
The heavy solid line is the most realistic yearly rate estimate.  The shaded
region represents the 5 to 95\% confidence interval arising from uncertainty
in predicted event rates \citep{Abadie:2010p10836}.}
\end{figure}

@@

Realizing advance detection of compact binary coalescences (\CBC{}s) will
require striking a balance between latency and throughput. \CBC{} searches
consist of banks of matched filters, or
cross-correlations between the data stream and a bank of nominal ``template''
signals.  There are many different implementations of matched filters, but most
have high throughput at the cost of high latency, or low latency at the cost of
low throughput.  The former are epitomized by the overlap-save algorithm~%
\citep{numerical-recipes-chapter-13} for frequency domain (\FD) convolution,
currently the preferred method in \GW{}
searches.  The most obvious example of the latter is the time domain
(\TD) convolution, which is latency-free.  However, its
computational complexity is quadratic in the length of the templates, so it is
prohibitively expensive for long templates.
